---
title: "EFA"
author: "Eleanor Hanna"
date: "April 30, 2018"
output: html_document
---

%%%%%%%% INITIAL DATA SCREENING %%%%%%%%

The impact of missing values was assessed in SPSS (since the options available in R would not extend to the number of variables in the data set).  The remaining data are here:
```{r}
setwd("Z:/Ellie/2016-2017/dissertation/incubator/aim1/manuscript/analysis/1_assumptions_met_for_EFA/2_outliers_and_normality/")
EFAdata <- read.csv("EFA_outlier_normality.csv",header=TRUE)
```

Univariate normality was assessed.
```{r}
library(MVN)
uniAnalysis <- mvn(EFAdata[,2:length(EFAdata)], univariateTest = "SW")
print(uniAnalysis$Descriptives$Skew)
print(uniAnalysis$Descriptives$Kurtosis)
```

Univariate distributions were inspected in order to identify items with skewness greater than |2| and/or kurtosis greater than |7| (see Curran et al., 1996).
```{r}
tooSkew <- which(uniAnalysis$Descriptives$Skew<(-2) | uniAnalysis$Descriptives$Skew>2)
tooKurt <- which(uniAnalysis$Descriptives$Kurtosis<(-7) | uniAnalysis$Descriptives$Kurtosis>7)
tooSkewVarbles <- c(uniAnalysis$univariateNormality$Variable)[tooSkew]
tooKurtVarbles <- c(uniAnalysis$univariateNormality$Variable)[tooKurt]
print(tooSkewVarbles)
print(tooKurtVarbles)
```

While no variables were kurtotic greater than |7|, two were skewed greater than |2|.

At the same time, I inspected variables for univariate outliers.
```{r}
outlierVarbles <- c()
numOutliers <- c()
for (varble in colnames(EFAdata[,2:length(EFAdata)])){
  zVarble <- (EFAdata[[varble]]-mean(EFAdata[[varble]]))/sd(EFAdata[[varble]])
  zOutliers <- which(zVarble>3.29 | zVarble<(-3.29))
  if (length(zOutliers) > 0) {
    outlierVarbles <- c(outlierVarbles, varble)
    numOutliers <- c(numOutliers, length(zOutliers))
    }
}
print(outlierVarbles)
print(numOutliers)
```

I wanted to record the skewness and kurtosis for variables that had univariate outliers.
```{r}
for (varble in outlierVarbles){
  print(varble)
  SK <- mvn(EFAdata[[varble]], univariateTest = "SW")
  print(SK$Descriptives$Skew)
  print(SK$Descriptives$Kurtosis)
}
```


In accordance with recommmendations from Tabachnick & Fidell 2013, I attempted transformations of these variables to address both the skewness and the outliers.  I began with the most conservative (reflect if negative skew, and square root).

D102...
```{r}
sqrtD102 <- sqrt(8-EFAdata[["D102"]])
SKsqrtD102 <- mvn(sqrtD102, univariateTest = "SW")
D102_z_sqrt <- ((sqrtD102-mean(sqrtD102))/sd(sqrtD102))
sqrtD102outliers <- which(D102_z_sqrt>3.29 | D102_z_sqrt<(-3.29))
print(SKsqrtD102$Descriptives$Skew)
print(SKsqrtD102$Descriptives$Kurtosis)
print(length(sqrtD102outliers))
```

D081...
```{r}
sqrtD081 <- sqrt(8-EFAdata[["D081"]])
SKsqrtD081 <- mvn(sqrtD081, univariateTest = "SW")
D081_z_sqrt <- ((sqrtD081-mean(sqrtD081))/sd(sqrtD081))
sqrtD081outliers <- which(D081_z_sqrt>3.29 | D081_z_sqrt<(-3.29))
print(SKsqrtD081$Descriptives$Skew)
print(SKsqrtD081$Descriptives$Kurtosis)
print(length(sqrtD081outliers))
```

D007...
```{r}
sqrtD007 <- sqrt(8-EFAdata[["D007"]])
SKsqrtD007 <- mvn(sqrtD007, univariateTest = "SW")
D007_z_sqrt <- ((sqrtD007-mean(sqrtD007))/sd(sqrtD007))
sqrtD007outliers <- which(D007_z_sqrt>3.29 | D007_z_sqrt<(-3.29))
print(SKsqrtD007$Descriptives$Skew)
print(SKsqrtD007$Descriptives$Kurtosis)
print(length(sqrtD007outliers))
```

D012...
```{r}
sqrtD012 <- sqrt(8-EFAdata[["D012"]])
SKsqrtD012 <- mvn(sqrtD012, univariateTest = "SW")
D012_z_sqrt <- ((sqrtD012-mean(sqrtD012))/sd(sqrtD012))
sqrtD012outliers <- which(D012_z_sqrt>3.29 | D012_z_sqrt<(-3.29))
print(SKsqrtD012$Descriptives$Skew)
print(SKsqrtD012$Descriptives$Kurtosis)
print(length(sqrtD012outliers))
```

D084...
```{r}
sqrtD084 <- sqrt(8-EFAdata[["D084"]])
SKsqrtD084 <- mvn(sqrtD084, univariateTest = "SW")
D084_z_sqrt <- ((sqrtD084-mean(sqrtD084))/sd(sqrtD084))
sqrtD084outliers <- which(D084_z_sqrt>3.29 | D084_z_sqrt<(-3.29))
print(SKsqrtD084$Descriptives$Skew)
print(SKsqrtD084$Descriptives$Kurtosis)
print(length(sqrtD084outliers))
```

D067...
```{r}
sqrtD067 <- sqrt(8-EFAdata[["D067"]])
SKsqrtD067 <- mvn(sqrtD067, univariateTest = "SW")
D067_z_sqrt <- ((sqrtD067-mean(sqrtD067))/sd(sqrtD067))
sqrtD067outliers <- which(D067_z_sqrt>3.29 | D067_z_sqrt<(-3.29))
print(SKsqrtD067$Descriptives$Skew)
print(SKsqrtD067$Descriptives$Kurtosis)
print(length(sqrtD067outliers))
```

D013...
```{r}
sqrtD013 <- sqrt(8-EFAdata[["D013"]])
SKsqrtD013 <- mvn(sqrtD013, univariateTest = "SW")
D013_z_sqrt <- ((sqrtD013-mean(sqrtD013))/sd(sqrtD013))
sqrtD013outliers <- which(D013_z_sqrt>3.29 | D013_z_sqrt<(-3.29))
print(SKsqrtD013$Descriptives$Skew)
print(SKsqrtD013$Descriptives$Kurtosis)
print(length(sqrtD013outliers))
```

D017...
```{r}
sqrtD017 <- sqrt(8-EFAdata[["D017"]])
SKsqrtD017 <- mvn(sqrtD017, univariateTest = "SW")
D017_z_sqrt <- ((sqrtD017-mean(sqrtD017))/sd(sqrtD017))
sqrtD017outliers <- which(D017_z_sqrt>3.29 | D017_z_sqrt<(-3.29))
print(SKsqrtD017$Descriptives$Skew)
print(SKsqrtD017$Descriptives$Kurtosis)
print(length(sqrtD017outliers))
```

D058...
```{r}
sqrtD058 <- sqrt(8-EFAdata[["D058"]])
SKsqrtD058 <- mvn(sqrtD058, univariateTest = "SW")
D058_z_sqrt <- ((sqrtD058-mean(sqrtD058))/sd(sqrtD058))
sqrtD058outliers <- which(D058_z_sqrt>3.29 | D058_z_sqrt<(-3.29))
print(SKsqrtD058$Descriptives$Skew)
print(SKsqrtD058$Descriptives$Kurtosis)
print(length(sqrtD058outliers))
```

D074...
```{r}
sqrtD074 <- sqrt(8-EFAdata[["D074"]])
SKsqrtD074 <- mvn(sqrtD074, univariateTest = "SW")
D074_z_sqrt <- ((sqrtD074-mean(sqrtD074))/sd(sqrtD074))
sqrtD074outliers <- which(D074_z_sqrt>3.29 | D074_z_sqrt<(-3.29))
print(SKsqrtD074$Descriptives$Skew)
print(SKsqrtD074$Descriptives$Kurtosis)
print(length(sqrtD074outliers))
```

If this did not reduce the skew and kurtosis to acceptable levels, and if there were still outliers, I reflected the scores (if necessary - which happened to be all of them) and implemented a log10 transform.

D102...
```{r}
logD102 <- log10(8-EFAdata[["D102"]])
SKlogD102 <- mvn(logD102, univariateTest = "SW")
D102_z_log <- ((logD102-mean(logD102))/sd(logD102))
logD102outliers <- which(D102_z_log>3.29 | D102_z_log<(-3.29))
print(SKlogD102$Descriptives$Skew) #empty??
print(SKlogD102$Descriptives$Kurtosis) #
print(length(logD102outliers))
```

D007...
```{r}
logD007 <- log10(8-EFAdata[["D007"]])
SKlogD007 <- mvn(logD007, univariateTest = "SW")
D007_z_log <- ((logD007-mean(logD007))/sd(logD007))
logD007outliers <- which(D007_z_log>3.29 | D007_z_log<(-3.29))
print(SKlogD007$Descriptives$Skew) #empty??
print(SKlogD007$Descriptives$Kurtosis) #
print(length(logD007outliers))
```

D084...
```{r}
logD084 <- log10(8-EFAdata[["D084"]])
SKlogD084 <- mvn(logD084, univariateTest = "SW")
D084_z_log <- ((logD084-mean(logD084))/sd(logD084))
logD084outliers <- which(D084_z_log>3.29 | D084_z_log<(-3.29))
print(SKlogD084$Descriptives$Skew) #empty??
print(SKlogD084$Descriptives$Kurtosis) #
print(length(logD084outliers))
```

D013...
```{r}
logD013 <- log10(8-EFAdata[["D013"]])
SKlogD013 <- mvn(logD013, univariateTest = "SW")
D013_z_log <- ((logD013-mean(logD013))/sd(logD013))
logD013outliers <- which(D013_z_log>3.29 | D013_z_log<(-3.29))
print(SKlogD013$Descriptives$Skew)
print(SKlogD013$Descriptives$Kurtosis)
print(length(logD013outliers))
```

The original nonnormal variables were removed...
```{r}
badItems = c('D007','D012','D013','D017','D058','D067','D074','D081','D084','D102')
for (badItem in badItems){
  badItemIndex <- which(colnames(EFAdata)==badItem)
  matA <- EFAdata[1:(badItemIndex-1)]
  matB <- EFAdata[(badItemIndex+1):length(EFAdata)]
  EFAdata <- cbind(matA,matB)
}
```

...and replaced with the transformed variables.
```{r}
EFAdata <- cbind(EFAdata,logD007,sqrtD012,logD013,sqrtD017,sqrtD058,sqrtD067,sqrtD074,sqrtD081,logD084,logD102)
```

Next, I assessed the transformed data set for multivariate outliers.
```{r}
library(psych)
multiChis <- outlier(EFAdata[,2:length(EFAdata)], plot=TRUE)
critChi <- 176.014 #https://mathcracker.com/chi-square-critical-values.php#results; right-tailed, df = 126
multiOutliers <- which(multiChis > critChi)
multiOutlierIDs <- EFAdata[["ResponseID"]][multiOutliers]
print(multiOutlierIDs)
```

I created a data set without the multivariate outliers:
```{r}
noMVOEFAdata <- EFAdata[-c(multiOutliers), ]
```

Then I assessed the multivariate normality of both datasets:
```{r}
mvn(EFAdata[,2:length(EFAdata)],mvnTest = "mardia",multivariatePlot="qq")
mvn(noMVOEFAdata[,2:length(EFAdata)],mvnTest = "mardia",multivariatePlot="qq")
```
Both datasets are multivariate nonnormal, so I decided to proceed with the factor analysis using MINRES.

As a final data cleaning step, I assessed the data set for multicollinearity and singularity.
```{r}
smc(EFAdata[,2:length(EFAdata)])
print(which(smc(EFAdata[,2:length(EFAdata)])>.9))
```

Tolerance for SMCs range up to about .99 on the low end (see T&F), and since none of these variables exceeded .90, I determined there was no multicollinearity or singularity present.


%%%%%%%% FACTORABILITY %%%%%%%%

I assessed the data set for sampling adequacy.
```{r}
sampAds <- KMO(EFAdata[,2:length(EFAdata)])
print(sampAds)
print(which.min(sampAds$MSAi))
```


%%%%%%%% NUMBER OF FACTORS %%%%%%%%

First I conducted an initial run using principal components extraction, in order to figure out the eigenvalues which are greater than one.
```{r}
eigenQuery <- principal(EFAdata[,2:length(EFAdata)], rotate="varimax")
jiffy <- which(eigenQuery$values > 1)
print(length(jiffy))
```
This is probably an unreasonable number of factors, and since the number of variables exceeds 40 and the communalities aren't that uniformly great, I did not use this value.

Next I conducted a parallel analysis to ascertain the number of factors to extract.
```{r}
parallel <- fa.parallel(EFAdata[,2:length(EFAdata)], fm = "minres", fa = 'fa')
```

I also used Velicer's Minimum Average Partial test and the Very Simple Structure criterion.
```{r}
vss(EFAdata[,2:length(EFAdata)], rotate = "oblimin", fm = "minres")
```
The VSS does not make much sense, because the item complexity is generally high (spoiler alert again).  The MAP recapitulates the parallel analysis.


%%%%%%%% THE ACTUAL EFA %%%%%%%%

The number of factors is either 7 or 8. After Costello & Osborne (2005,) I am running the analysis with the number of factors both below and above the suggestion of the scree plot, parallel and analysis, and MAP.
```{r}
library(GPArotation)
EFAminres6 <- fa(EFAdata[,2:length(EFAdata)], nfactors = 6, rotate = "oblimin", fm = "minres")
EFAminres7 <- fa(EFAdata[,2:length(EFAdata)], nfactors = 7, rotate = "oblimin", fm = "minres")
EFAminres8 <- fa(EFAdata[,2:length(EFAdata)], nfactors = 8, rotate = "oblimin", fm = "minres")
EFAminres9 <- fa(EFAdata[,2:length(EFAdata)], nfactors = 9, rotate = "oblimin", fm = "minres")
```

They were then inspected (in Excel, commenting out each analysis as needed).
```{r}
#print(EFAminres6)
#print(EFAminres7)
#print(EFAminres8)
print(EFAminres9)
```

The 7:9-factor solutions generally departed from simple structure pretty badly, getting factors where there were two or fewer non-complex variable loadings, so I dug down more into 5 and 6 by looking at them minus the items that didn't load.

The 7- and 9-factor analysis made a factor with no salient items; the 8-factor analysis made a factor with just 2 and a factor with just 3. The 6-factor analysis was the best.

If anyone was interested in whether this analysis survives if I remove the non-loading items, I conducted the 6-factor analysis with the items that didn't load.  This took 3 total iterations, including the first analysis.  The "badItems" vector is where I put the items that failed to load saliently onto any factor at any point in the process.
```{r}
reducedEFAdata <- EFAdata
badItems <- c('D002','D005','D031','D060','D071','D088','D122','D125','D126','D050')
for (badItem in badItems){
  badItemIndex <- which(colnames(reducedEFAdata)==badItem)
  if(badItemIndex==length(reducedEFAdata)){reducedEFAdata <- reducedEFAdata[1:length(reducedEFAdata)-1]}
  else{
    matA <- reducedEFAdata[1:(badItemIndex-1)]
    matB <- reducedEFAdata[(badItemIndex+1):length(reducedEFAdata)]
    reducedEFAdata <- cbind(matA,matB)
}}
EFAminres6red <- fa(reducedEFAdata[,2:length(reducedEFAdata)], nfactors=6, rotate="oblimin", fm="minres")
print(EFAminres6red)
```

